A preliminary constraint grammar for Russian

### 2015-03-30 2015-nodalida-ruscg/ruscg.tex
fixed ambiguity frequency table
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@59568 72bbbca6-d526-0410-a7d9-f06f51895060
замков etc.
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@59571 72bbbca6-d526-0410-a7d9-f06f51895060

\usepackage{times}
\usepackage{latexsym}
\usepackage{fixltx2e} %allows subscripts
%\usepackage{mathptmx}
%\usepackage{txfonts}
\usepackage{url}
\usepackage{alltt}
\usepackage{graphicx}
\usepackage[small,bf]{caption}
\special{papersize=210mm,297mm} % to avoid having to use "-t a4" with dvips 
%\setlength\titlebox{6.5cm}  % You can expand the title box if you really have to

@@

State-of-the-art morphological analysis in Russian is primarily based on
finite-state technology
\cite{Nozhov-03,Segalovich-03}.\footnote{Machine-learning approaches have also been successfully applied to Russian, most notably by 
\newcite{sharoff08lrec-mocky}.} Almost without exception, all large-scale 
morphological transducers of Russian are based on the
forward-looking \emph{Grammatical Dictionary of Russian} \cite{Zaliznjak-77}.
This dictionary gives fine-grained morphological specifications for more than
100~000 words, including inflectional endings, morphophonemic alternations, 
stress patterns, exceptions, and idiosyncratic collocations.
We base our work on the recently developed [anonymous] morphological transducer, which is also based on Zaliznjak's dictionary (Author, 2015).\footnote{The [anonymous] transducer is implemented using a two-level 
morphology \cite{Koskenniemi-84}, and can be compiled using either {\tt xfst}
\cite{Beesley.Karttunen-03} or {\tt hfst} \cite{hfst-11}} This finite-state transducer (FST) generates all possible 
morphosyntactic readings of each wordform, regardless of its frequency or probability. Because Russian is a relatively highly inflected language, broad coverage is important, but widespread homonymy leads to the generation of many spurious readings, as discussed in Section~\ref{ambiguity} below. Because of this, one of the foundational steps in Russian natural language processing is \rus{снятие омонимии} `morphosyntactic disambiguation'.

@@

\begin{table}
  \centering
  \begin{tabular}{l|rr}
    \hline
    \textbf{Type}  & \textbf{all tokens} & \textbf{ambig. tokens} \\
    \hline
    Intraparadigm. & 0.590                   & 0.909   \\
    Incongruent    & 0.277                   & 0.427   \\ 
    Congruent      & 0.120                   & 0.180    \\ 
    \hline
  \end{tabular}
  \caption{Proportional frequency of different types of morphosyntactic ambiguity in unrestricted text}
  \label{table:ambiguity}
\end{table}

% numbers not perfect, e.g. лицо should be 'cong.' but will turn up as 'intra.'

@@

\subsection{Morphological analyser}
\label{sec:morph}

The morphological transducer used in this study is described in Author (2015). 
It is primarily based on Zaliznjak's ``Grammatical dictionary of Russian'', 
including the 2001 version's appendix of proper nouns. It also includes approximately CITE Grishina and Lyashevskaya 2008. \url{http://dict.ruslang.ru/gram.php} .

@@

\ex. \label{ex:FSToutput} 
	\a. \rus{новый}\texttt{{\small <adj><m><nn><sg><nom>}}
	\b. \rus{автомат}\texttt{{\small <n><m><nn><sg><nom>}}
	\c. 
% det/prn (e.g. pos... их, я, мой)
% 'over'generation (e.g. all dets get all gen/nbr/case combinations)

@@

>>>>

\usepackage{times}
\usepackage{latexsym}
\usepackage{fixltx2e} %allows subscripts
\usepackage{tipa} %allows IPA symbols
\usepackage{linguex}
%\usepackage{mathptmx}
%\usepackage{txfonts}
\usepackage{url}
\usepackage{alltt}
\usepackage{graphicx}
\usepackage[small,bf]{caption}
\special{papersize=210mm,297mm} % to avoid having to use "-t a4" with dvips 
%\setlength\titlebox{6.5cm}  % You can expand the title box if you really have to

@@

State-of-the-art morphological analysis in Russian is primarily based on
finite-state technology
\cite{Nozhov-03,Segalovich-03}.\footnote{Machine-learning approaches have also been successfully applied to Russian, most notably by 
\newcite{sharoff08lrec-mocky}.} Almost without exception, all large-scale 
morphological transducers of Russian are based on the
forward-looking \emph{Grammatical Dictionary of Russian} \cite{Zaliznjak-77}.
This dictionary gives fine-grained morphological specifications for more than
100~000 words, including inflectional endings, morphophonemic alternations, 
stress patterns, exceptions, and idiosyncratic collocations.
We base our work on the recently developed [anonymous] morphological transducer, which is also based on Zaliznjak's dictionary (Author, 2015).\footnote{The [anonymous] transducer is implemented using a two-level 
morphology \cite{Koskenniemi-84}, and can be compiled using either {\tt xfst}
\cite{Beesley.Karttunen-03} or {\tt hfst} \cite{hfst-11}} This finite-state transducer (FST) generates all possible 
morphosyntactic readings of each wordform, regardless of its frequency or probability. Because Russian is a relatively highly inflected language, broad coverage is important, but widespread homonymy leads to the generation of many spurious readings, as discussed in Section~\ref{ambiguity} below. Because of this, one of the foundational steps in Russian natural language processing is homograph disambiguation.

@@

\begin{table}
  \centering
  \begin{tabular}{l|rr}
    \hline
    \textbf{Type}  & \textbf{all tokens} & \textbf{ambig. tokens} \\
    \hline
    Intraparadigm. & 59.0\%                   & 90.9\%   \\
    Incongruent    & 27.7\%                   & 42.7\%   \\ 
    Congruent      & 1.2\%                   & 1.8\%    \\ 
    \hline
  \end{tabular}
  \caption{Frequency of different types of morphosyntactic ambiguity in unrestricted text}
  \label{table:ambiguity}
\end{table}

Note that the percentages in Table~\ref{table:ambiguity} overlap, i.e. the readings of a given wordform
can have more than one kind of ambiguity. For example, the wordform \emph{zamkov} has the readings given in \ref{ex:zamkov}.

\ex. \label{ex:zamkov}
	 \a. \rus{замок}$^1$+N+Msc+Inan+Pl+Gen \label{ex:zamkovA}
	 \b. \rus{замок}$^2$+N+Msc+Inan+Pl+Gen \label{ex:zamkovB}
	 \c. \rus{замковый}+A+Msc+Sg+Pred      \label{ex:zamkovC}

The ambiguity between \ref{ex:zamkovA} and \ref{ex:zamkovB} is morphosyntactically congruent, and
the ambiguity between \ref{ex:zamkovA}/\ref{ex:zamkovB} and \ref{ex:zamkovC} is
morphosyntactically incongruent, so this wordform would be counted for both categories in
Table~\ref{table:ambiguity} above.

@@

\subsection{Morphological analyser}
\label{sec:morph}
%ROB

The morphological transducer used in this study is described in Author (2015). 
It is primarily based on Zaliznjak's \emph{Grammatical dictionary of Russian}, 
including the 2001 version's appendix of proper nouns. It also includes approximately CITE Grishina and Lyashevskaya 2008. \url{http://dict.ruslang.ru/gram.php} .

@@

\ex. \label{ex:FSToutput} 
	\a. \rus{новый}\texttt{{\small <adj><m><nn><sg><nom>}}
	\b. \rus{автомат}\texttt{{\small <n><m><nn><sg><nom>}}
	\c. 
% det/prn (e.g. pos... их, я, мой)
% 'over'generation (e.g. all dets get all gen/nbr/case combinations)
%^^^
%ROB

@@

The constraint grammar described in this paper has been applied to the task of automatic
word stress placement (Author, 2015). This task is especially relevant for 
Russian language learners, because vowels are pronounced differently depending on their position 
relative to stress position. For example, the word \emph{molokó} `milk' is pronounced 
/\textipa{m@l2kO}/, where each instance of the letter \emph{o} corresponds to a different vowel sound. 
Russian has complicated patterns of shifting stress, which are difficult for learners to master.
Almost 99\% of wordforms with ambiguous stress position can be disambiguated morphosyntactically,
so a constraint grammar can potentially resolve most stress ambiguity indirectly. The results of Author 2015 show that our constraint grammar overcomes about 42\% of the ambiguity relevant to stress
ambiguity in unrestricted text.
